---
layout: post
title:  "Rolling a Join"
date:   2016-09-12
comments: true
categories: data.table
---

```{r setup, include=FALSE}
library(data.table)
knitr::render_jekyll()
opts_knit$set(base.dir = '~/itsakettle.github.io/')
knitr::opts_chunk$set(echo = TRUE, fig.path='assets/img/rolling_joins/figure/')
```

The [data.table](https://cran.r-project.org/web/packages/data.table/index.html) package in R facilitates rolling joins. I've stumbled across the term a few times so I thought I should learn about it. My first Google search for `rolling join` yielded a mix of data topics...and other content that wasn't work safe. I carefully navigated to a few different pages and did some reading. I didn't understand what was going on though. So I created some toy examples of my own and with some tinkering I got to grips with the technique. 

# Nomenclature

According to the [data.table intro vignette](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.pdf) a *rolling join* is also called a *fast ordered* join or *last observation carried forward* join. The basic idea is that if there is no matching value for a row being looked up during the join, then the value from the closest record is matched in lieu. I guess each of the names kind of makes sense to me but I'm sticking with rolling join because otherwise the title of this post wouldn't work.

# Making up data

To start with I made up a couple of pretend data.tables to join together.

```{r dt-one}
  dt <- data.table(id=c(1, 1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2, 2, 2),
                 t=c(100, 200, 300, 400, 500, 600, 700, 800, 100, 200, 300, 400, 500, 600, 700, 800),
                 key=c('id', 't'))
  dt
```

The `id` column can be any arbitrary entity but here I've just used 1 and 2. The column `t` represents time, or something similar to that, and again I've just picked arbitrary values. `id` is set as the first key column and `t` is set as the second key column - the order will matter later.

Say we want to add another column, the colour associated with an `id` at time `t`.

```{r dt-two}
  dt.colour <- data.table(id=c(1, 2, 2),
                          t=c(300, 200, 600),
                          colour=c('blue', 'yellow', 'red'), key=c('id', 't'))
  dt.colour
```

So `id` 1 was associated with 'blue' at time 300 and `id` 2 was associated with 'yellow' at time 200 and 'red' at time 600. The keys are set the same as `dt`.

Let's join the two data.tables together to add the colours. So for each entry in `dt` we want to try and find any matching values in `dt.colour`. In data.table land this is done like this

```{r join1}
  dt.colour[dt,]
```

As expected the colour column has been added with values if possible and with NAs inserted where there is no match.

# Rolling

Let's set `roll=Inf` and see what happens

```{r join2}
  dt.colour[dt, , roll=Inf]
```

For rows that have no exact key match, the last value matching the first key is rolled forward. The last value is defined by the order of the final key column - in this case `t`. The `roll` argument is set to `Inf` to specify that there is no limit to how far a value can be rolled forward.

From here on I'm going to add the unrolled version of the colour column to `dt` so as the rolled result can be easily compared to non-rolled result.

```{r add-non-rolled-column}
  dt[, non.rolled.colour := dt.colour[dt, colour, roll=FALSE]]
  dt
```

# Equivalent arguments

Note that `roll=TRUE` and `roll=T` are equivalent to `roll=Inf`.

# Rolling backwards

Rolling can be reversed by setting `roll=-Inf`. Now values are rolled back instead of forward.

```{r join3}
  dt.colour[dt, , roll=-Inf]
```

# Limiting the roll

By setting roll to a number you can limit how far values are rolled forward or back. The value specified is relative to the second key column, in this case `t`. 


```{r join4}
  dt.colour[dt, , roll=100]
```

Since t increases in increments of 100 for each id, the colour is rolled forward one row. So with these datatables if `roll=99` then nothing will be rolled forward and the _result_ is the same as `roll=FALSE`.

```{r join5}
  dt.colour[dt, , roll=99]
```

What happens if roll isn't an integer? It seems to work as you would expect.

```{r join6}
  dt.colour[dt, , roll=100.1]
```

Here's an example with a higher value of `roll`.

```{r join7}
  dt.colour[dt, , roll=300]
```

Note again that when you specify a number for the `roll` argument it is how far a particular value can be rolled forward or back. It is worth pointing out that it can also be thought of as how far back a row missing a value is willing to look to find an alternative. This means you can use it to, say, find the most recent value within a certain window.

# Selective Roll
I couldn't figure out how to roll on some columns but not others. I don't think this is currently possible.

# Nearest Value

You can also set `roll=nearest`. Let's see what this does.

```{r roll-nearest}
  dt.colour[dt, , roll='nearest']
```

It looks like the nearest value, either forwards or backwards, is taken. You can see that there is a tie when `id=2` and `t=400` and in this case the value found by looking backwards seems to be taken over the one found by looking forwards. But what if you want to look only so far backwards and forwards?

# Rollends

There is also the `rollends` argument which by default is set to `c(TRUE, FALSE)` if `roll` is negative and `c(TRUE, FALSE)` if it is positive or zero (I suppose `roll=TRUE` is considered positive). So the first element of `rollends` is whether to roll backwards and the second element is whether to roll forwards. So by setting `rollends=c(TRUE, TRUE)` you can roll backwards and forwards whilst also specifying how far to look.

```{r rollends}
  dt.colour[dt, , roll=200, rollends=c(TRUE,TRUE)]
```

What happens when you set `rollends=c(FALSE, FALSE)`. I expected this would mean no rolling, but that doesn't seem to be the case. The documentation says 

> If rollends=FALSE the value of i must fall in a gap in x but not after the end or before the beginning of the data, for that group defined by all but the last join column.

First note that `rollends=FALSE` is the same as `rollends=c(FALSE, FALSE)` because the value is recycled. Next, in our case the `x` referred to is `dt.colours` and `i` is `dt`. So in order for a row in `dt` to find a match, its value of `t` must fall in between the values of `t` for two rows in `dt.colours` that have the same value of `id`, the first key column. Let's try it with `roll=Inf`. It might be a good idea to print out `dt.colours` as well for reference.

```{r rollends-false}
  dt.colour
  # Note roll=Inf
  dt.colour[dt, , roll=Inf, rollends=c(FALSE, FALSE)]
```

Only rows of `dt` that have a value of `t` in between values of `t` in dt.colour can roll previous values
forward. So when `id` is 1 there is no rolling since there is only one applicable row in `dt.colours` meaning nothing can be in between anything. However for `dt` equal 2 the rows with values of `t` from 200-500 are all in between the two entries with an `id` of 2 in `dt.colour`. So it seems that `rollends=FALSE` is a way of turning off rolling on the edges of the data.

# Multiple Grouping Columns

Everything used in this post has used two key columns `id` and `t` when performing rolling joins. However you can use as many as you like with the understanding that the last key column will always be used for rolling and the rest for defining groups. As a finish here's an example of this.

```{r dt-many-groups}
  dt <- data.table(id=c(1, 1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2, 2, 2),
                   id2=c(rep('a', 4), rep('b', 4), rep('a', 4), rep('b', 4)),
                 t=c(100, 200, 300, 400, 500, 600, 700, 800, 100, 200, 300, 400, 500, 600, 700, 800),
                 key=c('id', 'id2', 't'))
  dt
```

```{r dtcolour-many-groups}
  dt.colour <- data.table(id=c(1, 2, 2),
                          id2=c('b', 'b', 'b'),
                          t=c(300, 200, 600),
                          colour=c('blue', 'yellow', 'red'), key=c('id', 'id2', 't'))
  dt.colour
```

```{r many-column-roll}
# Add non rolled data for reference
dt[, non.rolled.colour := dt.colour[dt, colour, roll=FALSE]]
dt.colour[dt, , roll=TRUE]
```

